---
title: "Global Flows of Refugees"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    code_folding: hide
    incremental: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggformula)
library(networkD3)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,
                      message = FALSE,
                      fig.width = 6, 
                      fig.height = 7)

theme_set(theme_minimal(base_size = 16))

refuge <- read_csv('https://sldr.netlify.app/data/unhcr-refugee-data.csv',
                   skip = 14)
refuge2020 <- filter(refuge, 
                     Year == 2020)

sources <- refuge2020 %>%
  rename(origin = `Country of origin`,
         destination = `Country of asylum`,
         number = `Refugees under UNHCR’s mandate`) %>%
  group_by(origin) %>%
  mutate(origin = factor(origin)) %>%
  summarise(refugees_thousands = sum(number) / 1000) %>%
  mutate(origin = fct_reorder(origin,
                              refugees_thousands)) %>%
  ungroup() 

sources <- sources %>% 
  mutate(refugees_millions = refugees_thousands / 1000,
         refugees_percent = refugees_thousands / sum(pull(sources, refugees_thousands)) * 100) 

dest <- refuge2020 %>%
  rename(origin = `Country of origin`,
         destination = `Country of asylum`,
         number = `Refugees under UNHCR’s mandate`) %>%
  group_by(destination) %>%
  mutate(destination = factor(destination)) %>%
  summarise(refugees_thousands = sum(number) / 1000) %>%
  mutate(destination = fct_reorder(destination,
                              refugees_thousands)) %>%
  ungroup() 

dest <- dest %>% 
  mutate(refugees_millions = refugees_thousands / 1000,
         refugees_percent = refugees_thousands / sum(pull(dest, refugees_thousands)) * 100) 
```


## Who are Refugees?

*Note: you may have already watched these brief videos in preparation for class. If so, re-watching is optional.*

As you watch the video with your group, each person make note of:

- 3 key or surprising facts that grab your attention, and how they were visually illustrated. 
- Also make note of at least one question you have.

<iframe width="560" height="315" src="https://www.youtube.com/embed/96UmgY6E8n8" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

After watching, share your key facts and questions with your group. 

- Do you think the way the video visually presented certain facts affected which ones caught your eyes? 
- Can you answer each others' questions?

## The Resettlement Process

<iframe width="560" height="315" src="https://www.youtube.com/embed/mPu12hOoAu8" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

After watching the video with your group, talk briefly about what questions you have after seeing it. (If none come to mind, perhaps consider: where? how many?)

## Data

This exploration will let you explore further global movements of refugees, from 1951 to 2020.

The data come [from the UNHCR](https://www.unhcr.org/refugee-statistics/download/?url=Wrtv6K), downloaded in January 2021.

In R, we use the code below to read in the data:

```{r unhcr-data-in, exercise = TRUE}
refuge <- read_csv('https://sldr.netlify.app/data/unhcr-refugee-data.csv',
                   skip = 14)
```

### Get 2020 data
In addition to looking at data over time, we will also want to have a look at the current situation (well, in 2020).

We filter the data to keep only observations from the year 2020, storing the new dataset with the name `refuge2020`.

```{r keep-2020, exercise = TRUE}

```

```{r keep-2020-hint-1, eval = FALSE}
refuge2020 <- filter(refuge, 
                     year == ____)
```

```{r keep-2020-hint-2, eval = FALSE}
refuge2020 <- filter(refuge, 
                     Year == 2020)
```

## Objectives

Where did refugees come from, and where did they settle? How many? Your group will create and review several graphics to explore the data.

**One way to see it is that any graph tells a story. Another is that each graph answers a precise, concrete question. The best data analysts** *know* **what this question is as they create the graph, and make design choices to highlight the answer.**

As your group considers each graph, you may consider 3 key questions together:

1. What is the question the graph answers?
2. What is the answer?
3. What concrete features of the graph draw your focus to the answer? What, if anything, is distracting?

## Top Sources in 2020, Graph A

```{r top-sources, exercise = TRUE, fig.height=8, fig.width=7, echo = TRUE}
sources <- refuge2020 %>%
  rename(origin = `Country of origin`,
         destination = `Country of asylum`,
         number = `Refugees under UNHCR’s mandate`) %>%
  group_by(origin) %>%
  mutate(origin = factor(origin)) %>%
  summarise(refugees_thousands = sum(number) / 1000) %>%
  mutate(origin = fct_reorder(origin,
                              refugees_thousands)) %>%
  ungroup() 

sources <- sources %>% 
  mutate(refugees_millions = refugees_thousands / 1000,
         refugees_percent = refugees_thousands / sum(pull(sources, refugees_thousands)) * 100) 

gf_col(origin ~ refugees_millions,
       data = filter(sources, refugees_millions > 0.5)) %>%
  gf_labs(title = 'Refugees under UNHCR’s mandate',
          subtitle ='In millions, for countries with at least 500,000',
          x = 'Millions of people',
          y = '')
  
```

*For any teams interested in playing with the R code: can you figure out how to change the limit on which source countries are shown? Now it is ones with at least 500,000 refugees. Could you make it 250,000? Or 1,000,000?*

## Top sources in 2020, Plot B

```{r sources-prct, exercise = TRUE, fig.height=8, fig.width=7}
gf_col(origin ~ refugees_percent,
       data = filter(sources, refugees_percent > 1)) %>%
  gf_labs(title = 'Refugees under UNHCR’s mandate',
          subtitle ='omitting countries with < 1%',
          x = 'Percent',
          y = '')
  
```

*Is this graph different than the one before, other than showing percentages rather than numbers of people?  Which version is easier for you to make sense of? Which better emphasizes the fact that these are people, not just statistics? And can you think of other concrete changes to the graph that could emphasize that even more?*

## Destinations in 2020, Plot C

We've considered where refugees have come from; what about where they end up?

```{r top-dests, exercise = TRUE, fig.height=8, fig.width=8, echo = TRUE}
dest <- refuge2020 %>%
  rename(origin = `Country of origin`,
         destination = `Country of asylum`,
         number = `Refugees under UNHCR’s mandate`) %>%
  group_by(destination) %>%
  mutate(destination = factor(destination)) %>%
  summarise(refugees_thousands = sum(number) / 1000) %>%
  mutate(destination = fct_reorder(destination,
                              refugees_thousands)) %>%
  ungroup() 

dest <- dest %>% 
  mutate(refugees_millions = refugees_thousands / 1000,
         refugees_percent = refugees_thousands / sum(pull(dest, refugees_thousands)) * 100) 

gf_col(destination ~ refugees_millions,
       data = filter(dest, refugees_millions > 0.5 |
                       destination == 'United States of America')) %>%
  gf_labs(title = 'Refugees under UNHCR’s mandate',
          subtitle ='In millions; countries receiving at least 500,000 + USA',
          x = 'Millions of people',
          y = '')
  
```

*Any surprises here? Are there countries you did or did not expect to see?*

## Destinations in 2020, Plot D

```{r dest-prct, exercise = TRUE, fig.height=8, fig.width=8}
gf_col(destination ~ refugees_percent,
       data = filter(dest, 
                     refugees_percent > 1 |
                       destination == 'United States of America')) %>%
  gf_labs(title = 'Refugees under UNHCR’s mandate',
          subtitle ='including USA\notherwise omitting countries with < 1%',
          x = 'Percent',
          y = '')
  
```

*Compare the countries that are top sources and top destinations. Do you find any patterns? Can you imagine a way to show this data together, with both sources and destinations?*

## USA, over time

```{r usa-over-time, exercise = TRUE, fig.width = 8, fig.height = 3}
usa <- filter(refuge, `Country of asylum` == 'United States of America') %>%
  group_by(Year) %>%
  summarise(refugees = sum(`Refugees under UNHCR’s mandate`)) %>%
  ungroup()

gf_line(refugees / 1000 ~ Year,
        data = usa) %>%
  gf_labs(title = 'Refugees in the USA',
          y = 'Thousands of People') %>%
  plotly::ggplotly()
```

## USA, over time, different?

Consider a [graph from the Migration Policy Institute](https://public.tableau.com/views/RefugeeAdmissionstotheU_S_WRAPS1980-2020March/Dashboard1?:increment_view_count=no&:embed=y&:embed_code_version=3&:loadOrderID=0&:display_count=y&publish=yes&:origin=viz_share_link) that we might expect to look quite similar to our last one. What differences do you note? Can you identify possible reasons for the difference?

## USA and Everyone, over time

```{r all-over-time, exercise = TRUE, fig.width = 8, fig.height = 3}
USA <- filter(refuge, `Country of asylum` == 'United States of America') %>%
  group_by(Year) %>%
  summarise(usa = sum(`Refugees under UNHCR’s mandate`)) %>%
  ungroup()

all <- refuge %>%
  group_by(Year) %>%
  summarise(all = sum(`Refugees under UNHCR’s mandate`)) %>%
  ungroup() %>%
  left_join(y = USA) %>%
  pivot_longer(cols = c(usa, all),
               names_to = 'country',
               values_to = 'refugees')

gf_line(refugees/1000000 ~ Year,
        color = ~country,
        data = all) %>%
  gf_refine(scale_color_discrete('Destination')) %>%
  gf_labs(y = 'Millions of Refugees')
```

*Notice how the story of this plot is very different from the previous plots showing just US numbers over time!*

## Interaction

It is also possible to display this same dataset using a much more complex, and interactive visualization.

One example is the [data dashboard on the Migration Policy Institute website](https://www.migrationpolicy.org/programs/data-hub/charts/refugee-and-asylum-seeker-populations-country-origin-and-destination?width=1000&height=850&iframe=true).

*Investigate the dashboard for a few minutes and formulate two advantages and two disadvantages of this display, compared to our earlier simpler plots.*

## Conversation

*Have one group member take notes during this final conversation that you can hand in, and perhaps share with the whole class at the end (time permitting).*

How did this exercise affect your knowledge of global flows of refugees? Were there surprises, or things that really stuck in your mind?

Were there any discussions during this exercise that you can tie in to anything you've learned reading *Engaging God's World* so far?

What about links to other specific material from your section? Try to detail for your group-mates from the other section what that other material was all about, and how it relates.

## Learn More

Interested in digging deeper? There is so much more to learn! Check out the UNHCR's most recent (2019) report on [Global Trends in Forced Displacement](https://www.unhcr.org/5ee200e37.pdf).
