---
title: "US Incomes and Education"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(mosaic)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
incomes <- read_csv('https://sldr.netlify.com/data/incomes.csv') %>%
  mutate(education = forcats::fct_relevel(factor(education),
                                          "Less than 9th grade",
                                          "Some high school",
                                          "High school graduate",
                                          "Some college",
                                          "Associate degree",
                                          "Bachelor degree",
                                          "Master degree",
                                          "Professional degree",
                                          "Doctoral degree"))
```


## Data
Data on US median incomes was obtained from:

<from https://www2.census.gov/programs-surveys/cps/tables/time-series/historical-income-people/p16.xls>

The cleaned up data is hosted at: <https://sldr.netlify.com/data/incomes.csv>; read this data into R.

```{r data-in, exercise = TRUE}

```

```{r data-in-hint-1}
incomes <- read_csv( ___________ )
```

```{r data-in-hint-2}
incomes <- read_csv('https://sldr.netlify.com/data/incomes.csv')
```

## Trends
How does median income change over time?

```{r income-over-time, exercise = TRUE}

```


```{r income-over-time-hint-1}
gf_point(median_income ~ year, data = incomes)
```

## Education
Why do there seem to be so many different trend lines?  **Education** is one factor that strongly affects income. Color by `education` level to see:

```{r color-by-ed, exercise = TRUE}

```

```{r color-by-ed-hint-1}
gf_point(median_income ~ year, data = incomes, color = ~education)
```

## Sex
There still seem to be multiple lines for each education level. What is going on? What if we make the `shape` of the points depend on the worker's `sex`?

```{r income-by-sex, exercise = TRUE}

```


```{r income-by-sex-hint-1}
gf_point(median_income ~ year, data = incomes, 
         color = ~education,
         shape = ~sex)
```

**Now how would you describe the trends and patterns in US median incomes?**

*How might average (mean) incomes be different, and why does it probably make sense to use medians?*


## A Paradox
What is the trend in incomes between about 2000 and 2005, for almost all groups?

Many sources, however, report that US median incomes overall have **increased** in the same period.

*Both are actually true.*

How is this possible!?!
